---
title: The evolution of routes and paths in a dynamic environment
author:
  - name: Nicolas Gauthier
    email: Nicolas.Gauthier@asu.edu
    affiliation: Arizona State University
    footnote: Corresponding Author
address:
  - code: Arizona State University
    address: School of Human Evolution and Social Change, S. Caddy Mall, Tempe, AZ, Zip
abstract:

journal: "Journal of Archaeological Science"
date: "`r Sys.Date()`"
bibliography: mybibfile.bib
output: rticles::elsevier_article
---


```{r echo = FALSE, eval = FALSE}
# Install necessary packages if not already available (this chunk doesn't run by default).
install.packages('tidyverse')
install.packages('raster')
install.packages('NLMR')
install.packages('tidygraph')
install.packages('gifski')
install.packages('png')

# install the dev versions of these packages from github using devtools
install.packages('devtools')
devtools::install_github('thomasp85/ggraph')
devtools::install_github('thomasp85/gganimate')
devtools::install_github('thomasp85/patchwork')
```

```{r, echo = FALSE, message = FALSE}
library(raster)
library(tidyverse)
library(tidygraph)
library(ggraph)
library(NLMR)
library(gganimate)
library(patchwork)
```


# Introduction

Archaeology is embracing the form of constrained spatial interaction modeling developed by Wilson. These Boltzmann-Lotka-Volterra (BLV) style models use maximum entropy spatial interaction models to allocate flows between a spatially structured metapopulation, and Lotka-Volterra style consumer-resource equations to govern the growth of the populations. These equations are able to capture the dynamic feedbacks between settlements and the networks connecting them. Recent work in archaeology has expanded these models to allow the networks to further evolve, as routes that are more often used to connect important sites become themselves important, which in turn shapes the growth of settlements close to those routes. Past work has shown how stable routes and sets of routes can develop in mountainous topography, where physical constraints on movement are able to constrain the possible routes between settlements. 
  
Here, I extend this approach to examine how routes and paths -- the spatial networks that connect settlements -- evolve in response to patterns of environmental variability. Rather than leaving the carrying capacity of our population of settlements to remain fixed, we allow it to vary over space and in time. We examine how different patterns of spatio-temporal change lead to different settlement patterns and spatial network. We use a simple computatational modeling approach to facilitate a broad range of exploration, while maintaining interpretive clarity. We expect that different patterns, such as fixed oscillations, to lead to different stable patterns of spatial networks, whose dynamical behavior feeds back to influence settlement dynamics. In this way we seek to model the potential for "inertia" in settlement patterns, complicating the relationship between environmental forcing and social dynamics. Finally we allow for bidirectional feedbacks between human populations and the environment, exploring the potential for nonlinear social-ecological dynamics.

We distinguish here between "routes" and "paths". The former is a social construct, the latter are physical.


# Methods

Parameters for the model.

```{r}
n <- 100 # number of settlements
pop_start <- 150 # starting population per settlement
alpha <- 1.05 # superlinear returns to population size
beta <- 0.15 # distance decay parameter
```

We make a population of settlements. They all start with a population of `pop_start`, and are distributed randomly over a 100 x 100 space. 

```{r}
set.seed(3)
net <- tbl_graph(edges = expand.grid(from = 1:n, to = 1:n),
                 nodes = tibble(population = pop_start,
                                attractiveness = 1,
                                x = runif(100, max = 100),
                                y = runif(100, max = 100))) %E>%
  filter(!edge_is_loop()) %>%
  mutate(distance = sqrt((.N()$x[from] - .N()$x[to]) ^ 2 + 
                           (.N()$y[from] - .N()$y[to]) ^ 2)) %>%
  activate(nodes)

locations <- net %N>%
  select(x:y) %>%
  as_tibble
```


```{r}
net <- create_lattice(c(4, 4)) %N>%
  mutate(population = pop_start,
         attractiveness = 1,
         x = rep(c(20, 40, 60, 80), 4),
         y = rep(c(20, 40, 60, 80), each = 4)) %E>%
  mutate(distance = node_distance_to())

# need to figure out how to get full pairwise distance matrix and and to do that
```

```{r}
ggraph(net, x = x, y = y) + 
  geom_node_point(aes(size = population)) +
  geom_edge_link() +
  coord_equal() +
  theme_void()
```


```{r}
set.seed(100) # set seed for reproducability
pts <- tibble(x = sample(1:100, n, replace = TRUE),
              y = sample(1:100, n, replace = TRUE))
```

Then figure out the euclidean distances between the settlements and pull it all together into a `tbl_graph` object.

```{r}
settlements <- pts %>%
  dist(diag = TRUE, upper = TRUE) %>%
  as.matrix %>%
  as_tbl_graph %>%
  mutate(population = pop_start,
         attractiveness = population,
         x = pts$x,
         y = pts$y) %E>%
  rename(distance = weight) %>%
  filter(!edge_is_loop())
```

```{r echo = FALSE}
net %E>%
  group_by(from) %>% 
  top_n(-3, distance) %>% # retain the top 3 shortest edges
ggraph(node.positions = pts) +
  geom_node_point(aes(size = population)) +
  geom_edge_fan(aes(alpha = -1 *..index..), width = 1) +
  #scale_y_continuous(limits = c(1,100)) +
  #scale_x_continuous(limits = c(1,100)) +
  scale_size_area() +
  coord_equal() +
  ggtitle('Initial population distribution', 'k = 3 nearest neighbor network') +
  theme(legend.position="none") +
  theme_void()
```


## Landscape Model

We simulate different spatial patterns of environmental variability using the `NLMR` package.

```{r}
null <- matrix(0, nrow = 100, ncol = 100) %>% 
  raster(xmx = 100, ymx = 100)

uniform <- matrix(1, nrow = 100, ncol = 100) %>% 
  raster(xmx = 100, ymx = 100)

random <- nlm_random(100, 100)

distance_gradient <- 1- nlm_distancegradient(100, 100, origin = c(0, 10, 0, 10))

distance_gradient_center <- 1- nlm_distancegradient(100, 100, origin = c(49, 51, 49, 51))

oscillation <- distance_gradient * -1 + 1 - nlm_distancegradient(100, 100, origin = c(90, 100, 90, 100))

edge_gradient <- nlm_edgegradient(100, 100, direction = 0)

planar_gradient <- nlm_planargradient(100, 100, direction = 0)

grf_001 <- nlm_gaussianfield(100, 100, autocorr_range = 1, user_seed = 1000)
grf_010 <- nlm_gaussianfield(100, 100, autocorr_range = 10, user_seed = 1000)
grf_100 <- nlm_gaussianfield(100, 100, autocorr_range = 100, user_seed = 1000)


env <- brick(null, uniform, random, distance_gradient_center, edge_gradient, oscillation, grf_001, grf_010, grf_100)
```

```{r echo = FALSE}
env %>%
  as.data.frame(xy = TRUE, long = TRUE) %>%
  ggplot(aes(x, y)) +
  geom_raster(aes(fill = value)) +
  coord_equal() +
  facet_wrap(~layer) +
  theme_void() +
  scale_fill_viridis_c()
```


## Spatial Interaction Model

```{r interact}
interact <- function(net){
  net %E>%
    mutate(interaction_strength = .N()$attractiveness[to] ^ alpha * exp(-beta * distance))  %N>%
    mutate(outflow = population / centrality_degree(weights = interaction_strength, mode = 'out', loops = FALSE)) %E>%
    mutate(flow = .N()$outflow[from] * interaction_strength) %N>%
    mutate(inflow = centrality_degree(weights = flow, mode = 'in', loops = F),
           attractiveness = attractiveness + .01 * (inflow  - attractiveness),
           population = n * pop_start * attractiveness / sum(attractiveness))
}
```

```{r}
nystuen_dacey <- function(net){
  net %E>%
  group_by(from) %>%
  filter(flow == max(flow), .N()$population[from] < .N()$population[to]) %N>%
  mutate(terminal = node_is_sink()) %>%
  ungroup
}

# try using node_is_ functions to calculate instead
```

# Results

```{r message = FALSE}
sim_length <- 20 # number of time steps

sim <- accumulate(1:sim_length, ~interact(.x), .init = net) %>%
  .[2:sim_length] %>% # remove the initial state
  map(nystuen_dacey) %>%
  map(as_tibble) %>%
  bind_rows(.id = 'time') %>%
  mutate(time = as.numeric(time))
```


```{r, echo = FALSE}
ggplot(sim, aes(x, y, size = population)) +
  geom_point(aes(color = terminal)) +
  coord_equal() +
  theme_void() +
  labs(title = 'Year: {closest_state}') +
  transition_states(time)
```

Here are two sample references: @Feynman1963118 [@Dirac1953888].


# Discussion


# Conclusion


# References {#references .unnumbered}
