---
title: The evolution of routes and paths in a dynamic environment
author:
  - name: Nicolas Gauthier
    email: Nicolas.Gauthier@asu.edu
    affiliation: Arizona State University
    footnote: Corresponding Author
address:
  - code: Arizona State University
    address: School of Human Evolution and Social Change, S. Caddy Mall, Tempe, AZ, Zip
abstract:

journal: "Journal of Archaeological Science"
date: "`r Sys.Date()`"
bibliography: mybibfile.bib
output: rticles::elsevier_article
---


```{r echo = FALSE, eval = FALSE}
# Install necessary packages if not already available (not run by default)
install.packages(c('tidyverse', 'raster', 'NLMR', 'tidygraph', 'gifski', 'png', 'gdistance', 'furrr'))

# install the dev versions of these packages from github using devtools
install.packages('devtools')
devtools::install_github('thomasp85/ggraph')
devtools::install_github('thomasp85/gganimate')
devtools::install_github('thomasp85/patchwork')
devtools::install_github('ecohealthalliance/yenpathy')
```

```{r, echo = FALSE}
library(raster)
library(tidyverse)
library(tidygraph)
library(ggraph)
library(sf)
library(NLMR)
library(gganimate)
library(patchwork)
library(yenpathy)
library(gdistance)
library(furrr)
```


# Introduction

Paths are physical as well as social infrastructure. Hence the dynamics.


Archaeology is embracing the form of constrained spatial interaction modeling developed by Wilson. These Boltzmann-Lotka-Volterra (BLV) style models use maximum entropy spatial interaction models to allocate flows between a spatially structured metapopulation, and Lotka-Volterra style consumer-resource equations to govern the growth of the populations. These equations are able to capture the dynamic feedbacks between settlements and the networks connecting them. Recent work in archaeology has expanded these models to allow the networks to further evolve, as routes that are more often used to connect important sites become themselves important, which in turn shapes the growth of settlements close to those routes. Past work has shown how stable routes and sets of routes can develop in mountainous topography, where physical constraints on movement are able to constrain the possible routes between settlements. 
  
Here, I extend this approach to examine how routes and paths -- the spatial networks that connect settlements -- evolve in response to patterns of environmental variability. Rather than leaving the carrying capacity of our population of settlements to remain fixed, we allow it to vary over space and in time. We examine how different patterns of spatio-temporal change lead to different settlement patterns and spatial network. We use a simple computatational modeling approach to facilitate a broad range of exploration, while maintaining interpretive clarity. We expect that different patterns, such as fixed oscillations, to lead to different stable patterns of spatial networks, whose dynamical behavior feeds back to influence settlement dynamics. In this way we seek to model the potential for "inertia" in settlement patterns, complicating the relationship between environmental forcing and social dynamics. Finally we allow for bidirectional feedbacks between human populations and the environment, exploring the potential for nonlinear social-ecological dynamics.

We distinguish here between "routes" and "paths". The former is a social construct, the latter are physical.



  Sharing and exchange are critical for maintaining population in response to interannual climate variability, although this short term stability can come at the cost of long term environmental degradation \parencite{Janssen2010}. More restricted sharing rules can be effective \parencite{Hegmon1996}, but can also be sensitive to the asymmetric accrual of debts if one party's food supply is more volatile than anothers' \parencite{Crabtree2015}.
 
Social infrastructure interacts directly with physical infrastructure because social networks must map onto spatial networks. Metabolic costs, such as the energy expended producing and transporting food over space where transportation infrastructure is sparse, provide constraints on energy flows in exchange systems \parencite{Drennan1984}. In any particular case, the balance between these costs and the metabolic benefits of social interaction influences whether resources are moved in bulk to populations in need, or whether those populations move themselves to the available resources. The topology of spatial networks also constrains who can interact with whom, introducing bottlenecks and other structural flow constraints \parencite{Barthelemy2011SpatialNetworks}. Improvements to transportation infrastructure, such as roads and trails, decrease the effective distance between different settlements; failure to maintain these transportation networks increases the effective distance \parencite{McCall1985TheAfrica}.

Regional scale patterns of spatial interaction arise from the decisions of heterogeneous agents interacting with imperfect and incomplete information. They make decisions based on the perceived costs and benefits of interaction, as far as they are able to distinguish them. It is crucial to estimate distribution of each person's subjective predictions about the costs and benefits of social interaction, conditional on the information available to them about each potential destination.

Self organization ... settlemetn patterns represent spontaneous order in response to spatial variability. positive feedbacks

cite crema 2015 for vraying caryying capcity

Lotka-Volterra equations are used in ecology to model energy flows in a food web. More generally, these models can represent energy flows in any social-ecological system, such as an agricultural settlement extracting food from the surrounding hinterland. Start with a modified Lotka-Volterra consumer-resource system, in which an aggregate resource stock $X$ is harvested by a population of agriculturalists $N$. Assume $X$ is a renewable resource, so that in the absence of harvesting by $N$ it grows logistically with instantaneous growth rate $r$ and carrying capacity $K$. The rate that resources are harvested from $X$ and used to increase the population of $N$ is proportional to $HXN$, where $H$ is the constant \textit{per capita} harvest rate per unit of resource (for simplicity, assume that one unit of resource $X$ can be converted to one unit of population $N$). In the absence of resources, the population $N$ decays exponentially according to an intrinsic mortality rate $m$. The change in $X$ and $N$ over time can be described as a system of ordinary differential equations:

{\footnotesize  
  \setlength{\abovedisplayskip}{1pt}
  \setlength{\belowdisplayskip}{\abovedisplayskip}
  \setlength{\abovedisplayshortskip}{0pt}
  \setlength{\belowdisplayshortskip}{1pt}
\begin{equation}
    \begin{split}
        \dot{X} &= rX \left(1 - \frac{X}{K} \right) - HXN\\
        \dot{N} &= HXN - mN.
    \end{split}
\end{equation}}

After \textcite{Qubbaj2014}, this simple model can be extended into a social-ecological network of multiple interconnected consumer-resource systems. First, disaggregate $X$ and $N$ into $X_i$ and $N_j$, representing the resource at location $i$ and the population at location $j$. Then, replace the constant harvest rate $H$ with an adjacency matrix $\mathbf{H}$, such that population $N_j$ harvests resource $X_i$ at rate $H_{ij}$. If $X_i$ and $N_j$ are not connected on the network, then $H_{ij} = 0$. The proposed framework will relax two simplifying assumptions in this basic formulation: the lack of resource and population flows between settlements and the time-invariant intrinsic growth rate and carrying capacity of the resource.

\subsection{Geography: Exchange Networks and Spatial Interaction Models}
First, allow goods and people to flow between populations in $N$ via an adjacency matrix $\mathbf{E}$, representing a (social) food-exchange network embedded in space. $\mathbf{E}$ is derived from an entropy-maximizing spatial interaction model, which estimates the intensity of interaction between locations as a function of distance and a measure of their mutual "attractiveness" \parencite{Wilson2011EntropyModelling}. Entropy maximization is a means of making unbiased estimates of the most likely distribution of large-scale properties of a system (in this case a spatial network), making the fewest possible assumptions about micro-scale dynamics (social networks) \parencite{Presse2013PrinciplesPhysics}. Models that use entropy maximization to estimate the "fast" flow dynamics and consumer-resource equations such as (1) to represent the ``slow'' settlement dynamics are known as Boltzmann-Lotka-Volterra models \parencite{Wilson2006EcologicalTheory,Wilson2008BoltzmannSystems}, and useful for working with the limited information inherent to the archaeological record \parencite{Bevan2013ModelsEvidence,Davies2014ApplicationAges,Altaweel2015EvaluatingMaximization}.

The distribution of flows in $\mathbf{E}$ with the maximum entropy is determined by maximizing the flow functions subject to simple self-consistency constraints, using the method of Lagrange multipliers \parencite{Wilson2011EntropyModelling}. The maximum entropy solution representing $E_{ij}$, the flow of resources from site $i$ to site $j$, is:

{\footnotesize  
  \setlength{\abovedisplayskip}{1pt}
  \setlength{\belowdisplayskip}{\abovedisplayskip}
  \setlength{\abovedisplayshortskip}{0pt}
  \setlength{\belowdisplayshortskip}{1pt}
\begin{equation}
    E_{ij} = \frac{O_i W_j^\alpha e^{-\beta c_{ij}}}{\sum_kW_k^\alpha e^{-\beta c_{ik}}},
\end{equation}} where $O_i$ is the total outflows from site $i$, $W$ represents a site's attractiveness to site $i$, and $c$ is some function of distance. $\alpha$ and $\beta$ are parameters derived from the Lagrange multipliers; $\beta$ is particularly relevant here as it determines the impact of distance on flow intensity, and can be varied to represent the different costs of moving food versus people across an exchange network. Here, a site's attractiveness $W$ is its \textit{per capita} food supply, $\frac{N}{X}$. This differs from the standard formulation used in geography and urban studies, where a settlement's attractiveness is only a function of its size \parencite{Evans2011InteractionsModels}. Linking interaction intensity to food supply in this way allows the model to incorporate the influence of climate variability directly.

# Methods

We start with two basic entities, settlements and their hinterlands. Together these make up a simple consumer-resource system, with a population of humans in the settlement consuming agricultural resources from the hinterland. For simplicity, we assume the hinterland produces a fixed amount of food each year, varying only due to exogenous environmental factors (rainfall). The population in a settlement grows logistically, with a carrying capacity defined by the resources in its hinterland.

```{r}
logistic_growth <- function(x, k, r = 0.015){
  x * (1 + r * (1 - (x / k)))
}
```

```{r, echo = FALSE}
accumulate(rep(150, 500), logistic_growth, .init = 15) %>%
  tibble(time = 0:500, population = .) %>%
  ggplot(aes(time, population)) +
  geom_line(size = 1.2) +
  geom_hline(yintercept = 150, color = 'red', linetype = 2) +
  ggtitle('Settlement population over 500 years', 'Carrying capacity in red') +
  theme_classic()
```

A settlement is not limited to extracting food from its immediate hinterland, but rather can draw on resources from multiple hinterlands at greater distances. The amount of resources harvested from each hinterland is a function of the distance from the settlement to that hinterland. If distance played no role, we could simply add together all the discrete hinterland objects into an aggregate hinterland and recalculate the carrying capacity. Adding in this distance effect reduces this, limiting the areal extent of resource extraction from a site.
well maybe

The population of a settlement is free to move between other settlements. 


Parameters for the model.

```{r}
pop_start <- 100 # starting population per settlement
#alpha <- 1.05#1.05 # superlinear returns to population size
#beta <- 0.15 # distance decay parameter
```


We make a population of settlements. They all start with a population of `pop_start`, and are distributed evenly over a 100 x 100 space. We define the radius of the hexagonal cell as 5, setting a limit of 5km from a settlement as that settlement's "hinterland."

Why hexagons (birch et al 2007). Edge effects of the whole hexagonal domainare slo decreased
```{r}
radius <- 5

hex <- c(155, 13, 55, 13, 5, 100, 55, 187, 155, 187, 205, 100, 155, 13) %>%
  matrix(nrow = 7, byrow = 2) %>%
  list %>%
  st_polygon %>% 
  st_make_grid(cellsize = radius * 2, square = FALSE) %>%
  st_sf

n <- nrow(hex)

pts <- hex %>%
  st_centroid %>%
  st_coordinates
```

Now we create the physical network of paths connecting all neighboring settlements. 
```{r}
paths <- hex %>% 
  st_is_within_distance(dist = 1) %>%
  as.matrix %>% 
  `*`(1) %>% 
  as_tbl_graph() %E>%
  filter( from <= to) %>%#!edge_is_loop(), # or implement phi parameter instead
  mutate(distance = if_else(edge_is_loop(), radius, weight * radius * 2),
         effective_distance = distance,
         traffic = 0,
         total_traffic = 0) %>%
  select(-weight) %>%
  convert(to_undirected) %N>%
  mutate(x = pts[,1],
         y = pts[,2],
         food = 200,
         population = pop_start,
         attractiveness = 1, 
         edge = centrality_degree() < 6)
```

```{r, echo = FALSE}
ggraph(paths) + 
  geom_edge_link(alpha = .3) +
    geom_node_point() +
  coord_equal() +
  theme_void()
```


## Landscape Model

We simulate different spatial patterns of environmental variability using the `NLMR` package.

```{r}
null <- matrix(0, nrow = 210, ncol = 210) %>% 
  raster(xmx = 210, ymx = 210)
uniform <- matrix(1, nrow = 210, ncol = 210) %>% 
  raster(xmx = 210, ymx = 210)
#random <- nlm_random(210, 210)
distance_gradient <- 1- nlm_distancegradient(210, 210, origin = c(0, 10, 0, 10))
distance_gradient_center <- 1- nlm_distancegradient(210, 210, origin = c(105, 105, 105, 105))
oscillation <- distance_gradient * -1 + 1 - nlm_distancegradient(210, 210, origin = c(200, 210, 200, 210))
edge_gradient <- nlm_edgegradient(210, 210, direction = 0)
planar_gradient <- nlm_planargradient(210, 210, direction = 0)

grf_010 <- nlm_gaussianfield(210, 210, autocorr_range = 10, user_seed = 1000)
grf_100 <- nlm_gaussianfield(210, 210, autocorr_range = 100, user_seed = 1000)
grf_1000 <- nlm_gaussianfield(210, 210, autocorr_range = 1000, user_seed = 1000)
```


```{r}
env_names <- c('null', 'uniform','distance_gradient_center', 'edge_gradient', 'oscillation', 'grf_010', 'grf_100', 'grf_1000')
env <- brick(null, uniform, distance_gradient_center, edge_gradient, oscillation, grf_010, grf_100, grf_1000) %>%
  `names<-`(env_names) %>%
  raster::extract(hex, fun = mean, na.rm = TRUE, df = TRUE) %>%
  bind_cols(hex, .)
```

```{r echo = FALSE}
devtools::install_github("thomasp85/scico")

library(scico)
env %>%
  gather(variable, value, null:grf_1000) %>%
  mutate(variable = fct_relevel(variable, env_names)) %>%
  filter(!(variable %in% c('null', 'uniform', 'random'))) %>%
  ggplot() +
  geom_sf(aes(fill = value), lwd = .1) +
  facet_wrap(~variable) +
  theme_void() +
  scale_fill_scico(palette = 'roma', direction = -1) +
  coord_sf(datum = NA)
```

## Spatial Interaction Model

```{r interact}
interact <- function(net, alpha = 1.05, beta = 0.2){
  net %E>%
    mutate(utility = .N()$attractiveness[to] ^ alpha * exp(-beta * effective_distance))  %N>%
    mutate(outflow = population / centrality_degree(weights = utility,
                                                    mode = 'out', loops = FALSE)) %E>%
    mutate(flow = .N()$outflow[from] * utility) %>%
    select(-utility) %N>%
    select(-outflow)
  }

grow <- function(net, k = 1, epsilon = 0.1, G = 0.1){
  net %N>%
    mutate(inflow = centrality_degree(weights = flow, mode = 'in', loops = FALSE),
           inflow = if_else(edge == TRUE, inflow + inflow * G, inflow),
           attractiveness = attractiveness + epsilon * (inflow - k * attractiveness),
           population = n * pop_start * attractiveness / sum(attractiveness)) %>%
    select(-inflow)
}

# k convert units of flow to units of attractiveness. should be 1? we're interpreting units of flow as people, so probably not.... maybe we can say that a 1 of the climate variable is the amount of rain to produce enough food to feed xx people, and rescale from there?
```

```{r}
run_sim <- function(net, alpha = 1.15, beta_food = 0.15, beta_people = 0.25, nu = 0.05, k = 1, epsilon = 0.01, G = 0, a = 0.2, b = .0005){
  net %E>%
    #get_distance  %E>%
    # interact
    mutate(trade_utility = .N()$population[to] ^ alpha * exp(-beta_food * effective_distance))  %N>%
    mutate(trade_outflow = food / centrality_degree(weights = trade_utility, mode = 'out', loops = TRUE)) %E>%
    mutate(trade_flow = .N()$trade_outflow[from] * trade_utility) %N>% 
    mutate(trade_inflow = centrality_degree(weights = trade_flow, mode = 'in', loops = TRUE)) %E>%
    mutate(migrant_utility = .N()$trade_inflow[to] * .N()$population[to] ^ (alpha - 1) * effective_distance ^ (-beta_people)) %N>%
    mutate(migrant_outflow = nu * population / centrality_degree(weights = migrant_utility, mode = 'out', loops = FALSE)) %E>%
    mutate(migrant_flow = .N()$migrant_outflow[from] * migrant_utility) %N>%        
    mutate(migrant_inflow = centrality_degree(weights = migrant_flow, mode = 'in', loops = FALSE),
           #inflow = if_else(edge == TRUE, inflow + inflow * G, inflow),
           population = population + epsilon * (trade_inflow - k * population) - nu * population + migrant_inflow))

  #maintain(tmp, net, a, b)
}
```

```{r}
get_distance <- function(net){
  node_data <- as_tibble(net, 'nodes')
  
  dist_weights <- as_tibble(net, 'edges')$effective_distance
  
  net %>%
    igraph::distances(weights = dist_weights) %>% # needs weight column
    replace(. == 0, 999) %>% # replace 0 values with 999 temporarily
    as_tbl_graph(directed = TRUE) %>%
    mutate(!!!node_data) %E>%
    mutate(weight = if_else(weight == 999, 0, weight)) %>%
    rename(effective_distance = weight)
}
```



```{r}
maintain <- function(routes, paths, a, b){
  pop <- as_tibble(routes, 'nodes') %>% pull(population)
  act <- as_tibble(routes, 'nodes') %>% pull(attractiveness)

  tmp1 <- as_tibble(routes, 'edges') %>%
    group_by(from) %>%
      nest %>%
    mutate(path = map(from, get_paths, net = paths)) %>%
    select(from, path) %>%
    unnest(cols = c(path)) %>%
    mutate(to = rep(1:n, times = n)) %>% 
    left_join(as_tibble(routes, 'edges'), .) %>%
    mutate(len = map_dbl(path, length))
  
  tmp2 <- tibble(path =  unlist(tmp1$path),
           migrant_flow = rep(tmp1$migrant_flow, times = tmp1$len),
           trade_flow = rep(tmp1$trade_flow, times = tmp1$len)) %>%
        group_by(path) %>%
        summarise(trade_flow = sum(trade_flow),
                  migrant_flow = sum(migrant_flow))

  paths %E>%
    select(-traffic) %>%
    left_join(tmp2, by = c('.tidygraph_edge_index' = 'path')) %N>%
  # mutate(total_traffic = traffic + total_traffic,
   #        effective_distance = distance * improve_path(total_traffic, a = a, b = b)) %N>%
    mutate(population = pop,
           attractiveness = act)
}

get_paths <- function(from, net){
    
  dist_weights <- paths %E>%
    as_tibble %>%
    pull(effective_distance)
  
  shortest_paths(net, from, weights = dist_weights, output = 'epath') %>%
    .$epath %>% 
    map(as.numeric)
}

improve_path <- function(traffic, a, b){
  (1 - a) * exp(-b * traffic) + a
}
```

So instead of just this have a two sector model. flow of food from hinterland to centers ... this effects their carrying capacity. then flow of people between centers based on carryin capacity and size ... this is the qubbaj et al model with spatial interaction terms. as a first line, we assume that people only draw food from their direct hinterland, so instead we are looking at the flow of people between  the settlements based on food availability.

so does this mean we can vary k or what?

or have it be inflow + local carrying capacity  - k * attractiveness?

so it seems like we want to incorporate carrying capcaity into the population dynamics, not the interaction dynamics. that is the caryring capacity doesn't enter directly into the utilityes, but rather influences the settlement sizes which in turn effects the utilities. hmmmm not sure about this

maybe this is all just to day do the welfare modulated diffuse thing?

or have attractiveness based on food ratio, but the evolution rule only changes population endogenously and caryring capacity is left exogenous

maybe the phi thing from crema 2014 and 2015, with ratio of population to carrying capacity if above carrying capcity

I think there's still room for welfare modulated diffusion here

maybe have another epsioln development thing for migrant with a migration rate like .05


OOOOOOOH trade inflow is the attractiveness term for migrant flow


```{r}
nystuen_dacey <- function(net){
  net %E>%
  group_by(from) %>%
  filter(trade_flow == max(trade_flow)) %>%
  ungroup %>%
  filter(.N()$population[from] < .N()$population[to]) %N>%
  mutate(terminal = node_is_sink()) %>%
  convert(to_undirected) 
}

# try using node_is_ functions to calculate instead
```

```{r}
library(profvis)

profvis(run_sim(get_distances(paths)))
```


# Results

Maybe try running with all environmentalattreativeness and no population feedback, but instead with the growing roads?

or disaggregate so to attractivenesses are carrying capacity and size relative to carrying capacity?

```{r message = FALSE}
sim_length <- 1000 # number of time steps

sim <- accumulate(1:sim_length, ~run_sim(.x), .init = paths %>% get_distance) %>%
  .[2:sim_length]# remove the initial state
```

```{r}
qplot(seq(.1, 10, by = .1),2 *  sin(seq(.1, 10, by = .1)), geom = 'line')
```


```{r}
sim <- paths %N>%
  #mutate(food = food * env$edge_gradient) %>%
    mutate(population = population * env$edge_gradient) %>%
  get_distance() %>%
  accumulate(1:sim_length, ~run_sim(.x), .init = .) %>%
  .[2:sim_length]# remove the initial state
```

```{r, echo = FALSE}
anim <- sim %>%
   # map(nystuen_dacey) %>%
   map(as_tibble, 'nodes') %>%
  bind_rows(.id = 'time') %>%
  mutate(time = as.numeric(time)) %>%
  filter(time %% 10 == 0) %>%
ggplot(aes(x, y, size = population)) +
  geom_point()+#aes(color = terminal)) +
  coord_equal() +
  theme_void() +
  labs(title = 'Year: {closest_state}') +
  transition_states(time)

animate(anim, nframes = 199)
```

```{r}
sim[[sim_length - 1]] %>%
  as_tibble('nodes') %>%
  qplot(x = log(population),data = .,  geom = 'density')
sim[[sim_length - 1]]

sim %>%
  map(as_tibble, 'nodes') %>%
  bind_rows(.id = 'time') %>%
  mutate(time= as.numeric(time)) %>%
  ggplot(aes(time, population, group = .tidygraph_node_index)) +
  geom_line(alpha = .2) +
  theme_classic()


test <- sim[[sim_length - 1]] %E>%
  maintain(paths, a = 1, b = 0) %E>%
  arrange(trade) 
 # nystuen_dacey %>%

test %>%
ggraph(x = x, y = y) + 
  geom_edge_fan(aes(color = log(trade)), width = 1.2) +
   geom_node_point(aes(size = population)) +#, color = terminal)) +
  #scale_color_manual(values = c('black', 'red')) +
  scale_edge_color_viridis() +
  coord_equal() +
  theme_void()

sim[[sim_length - 1]] %E>% 
  nystuen_dacey() %>%
  ggraph(x = x, y = y) +
  geom_edge_link() +
  geom_node_point(aes(size = population)) +
  coord_equal() +
  theme_void()
```
The outer ring network reminds us of central place theory (Openshaw and Veneris 2003)


So generally we see more settlements with increasing beta. With increasing alpha we get a more defined heirarchy of site sizes.

Past a beta of .4 or whatever, the patterns don't change. this is because at this point, the distance decay function attentuates at or below the baseline 10km distance between settlements. Important to not here that the interperation of the value of beta depends on the characteristic length scale.
```{r}
plan(multisession)

param_sweep <- expand.grid(alpha = seq(1, 1.5, by = 0.05),
       beta = seq(0, 0.5, by = 0.05 )) %>%
  mutate(sim = future_map2(alpha, beta, function(x, y) reduce(1:sim_length, ~run_sim(., alpha = x, beta = y, G = 0, a = 1), .init = paths), .progress = TRUE))


param_sweep[[1,3]]

saveRDS(param_sweep, 'param_sweep')
param_sweep <- readRDS('param_sweep')
test <- param_sweep %>%
  mutate(dat = map(sim, as_tibble, 'nodes')) %>%
  select(-sim) %>%
  unnest(cols = c(dat))
```

"jumps" in the parameter space (clarke and wilson 1983 or 1985, regional science). basically the relationships between alpha and beta are consistent for this

```{r}
test %>%
  filter(population, beta > 0) %>%
  ggplot(aes(log(population))) +
  geom_density(fill = 'black') +
  facet_grid(beta ~ alpha) +
  theme_classic()
```

```{r}
test %>%
  arrange(population) %>%
  filter(population > 1, 
         beta > 0) %>%
  ggplot(aes(x, y, size = population)) +
  geom_point()+
  scale_size_area() +
  coord_equal() +
  theme_void() +
  facet_grid(beta ~ alpha)


hex %>%
  mutate(x = pts[,1],
         y = pts[,2]) %>%
  left_join(test, .) %>%
    filter(beta > 0 & beta < 0.4,
         alpha < 1.4) %>%
  ggplot() +
  geom_sf(aes(fill = log(population)), color = 'white', lwd = .1) +
  scale_fill_viridis_c() +
  coord_sf(datum = NA) +
  theme_void() +
  facet_grid(beta ~ alpha)


hex %>%
  mutate(x = pts[,1],
         y = pts[,2]) %>%
  left_join(test, .) %>%
    filter(population > 1,beta > 0 & beta < 0.4,
         alpha < 1.4) %>%
  ggplot() +
  geom_sf(aes(fill = log(population))) +
  scale_fill_viridis_c() +
  coord_sf(datum = NA) +
  theme_void() +
  facet_grid(beta ~ alpha)

  ggsave('hex.png', width = 12, height = 8)
  
```


```{r}
test2 <- param_sweep %>%
  mutate(sim = future_map(sim, ~maintain(., paths, a = 1, b = 0), .progress = TRUE))
  
test2 %>%
  .[[121, 3]] %E>%
  arrange(traffic) %>%
ggraph(x = x,  y = y) +
  geom_edge_link(aes(color = log(traffic)), width = 1.2) +
  scale_edge_color_viridis() +
  coord_equal() +
  theme_void()
  
test %>%
  arrange(population) %>%
  filter(population > 1, 
         beta > 0 & beta < 0.4,
         alpha < 1.4) %>%
  ggplot(aes(x, y, size = population)) +
  geom_point()+
  scale_size_area() +
  coord_equal() +
  theme_void() +
  facet_grid(beta ~ alpha)
```
run a clustering algorithm on the outputs to determine ?regions
```{r}
test3 <- param_sweep %>%
  mutate(simsim = map(sim, ~filter(., population >= 1)),
         sim = map(sim, ~mutate(., group = group_components())),
         dat = map(sim, as_tibble, 'nodes')) %>%
  select(-sim) %>%
  unnest(cols = c(dat))


param_sweep[[120,3]] %>%
  filter(population >= 1) %>%
  mutate(group = group_components())

test3 %>%
  filter(beta > 0) %>%
  group_by(alpha, beta, group) %>%
  summarise(n_cells = n()) %>%
  ggplot(aes(n_cells)) +
  geom_histogram() +
  facet_grid(alpha ~ beta)
```


Here are two sample references: @Feynman1963118 [@Dirac1953888].


# Discussion

Edge effects -- different ways of handling them. Sidestepped here by the shear size of the domain. hexagons have low perimeter to area ratio

Next, allow climate variability to influence resource dynamics. Assume a water-limited environment, so that the growth rate and carrying capacity of the resource represents the limitation of soil moisture on resource growth and abundance. The static parameters $r$ and $K$ are replaced with an unknown function $f\left(X_i, Q_i\right)$, representing logistic growth dependant on soil moisture $Q$ -- such functions are common in ecology \parencite[e.g., see][]{Ursino2007ModelingProcesses} but are kept generic here to enable application in variable ecohydrological contexts. As with $\mathbf{H}$ and $\mathbf{E}$ above, $Q$ can be replaced by an $I \times I$ adjacency matrix $\mathbf{Q}$, representing the pairwise directed moisture fluxes between all $X_i$ resource systems. This moisture-recycling network can be generated synthetically by the modeler, to explore different environmental scenarios, or estimated empirically from gridded observations, climate model outputs, or paleo-reanalysis data.

Empirical estimates of $\mathbf{Q}$ can be derived using functional network analysis, a method for inferring latent connectivity by comparing time series measured at every element of a complex system \parencite{Donges2015UnifiedPackage}. The elements of an empirical $\mathbf{Q}$ correspond to the \textit{transfer entropy} between time series of precipitation minus evapotranspiration (i.e. the land-atmosphere moisture flux) at each grid cell of a climate model. Transfer entropy is an information-theoretic measure analogous to standard measures of correlation, but extended to capture causal dependencies between nonlinear time series \parencite{Runge2012EscapingEntropy}. As such, a functional network representation of climate variability is conceptually and mathematically similar to the empirical orthogonal functions used in the American Southwest case but allows for more direct representations of the circulation dynamics leading to those patterns \parencite{Donges2015}. 

# Conclusion


# References {#references .unnumbered}

```{r}
knitr::knit_exit()
```


Experimenting with gdistance
```{r}
env <- matrix(1, nrow = 100, ncol = 100) %>% 
  raster(xmx = 100, ymx = 100)

euc_dist <- function(x1, x2, y1, y2){
  sqrt((x1 - x2) ^ 2 + (y1 - y2) ^2)
}

trans <- transition(env, transitionFunction = mean, directions = 8) %>%
  geoCorrection(type = 'r')

set.seed(1000)
settlements <- tbl_graph(edges = expand.grid(from = 1:n, to = 1:n),
                 nodes = tibble(population = pop_start,
                                attractiveness = 1,
                                x = sample(1:100, n, replace = TRUE),
                                y = sample(1:100, n, replace = TRUE))) %E>%
  mutate(euclidean = euc_dist(.N()$x[from], .N()$x[to], 
                             .N()$y[from], .N()$y[to])) %>%
  filter(!edge_is_loop())

sites <- settlements %N>%
  as_tibble() %>%
  select(x:y) %>%
  as.matrix

accCost(trans, sites) %>% plot

routes <- commuteDistance(trans, sites) %>%
  as.matrix %>%
  as_tbl_graph %E>%
  as_tibble %>%
  left_join(settlements, .) 


routes %>%
  as_tibble() %>%
  ggplot(aes(euclidean, weight)) + geom_point()

passage(trans, sites[2,], sites[1,], theta=5) %>% plot

plan(multicore, workers = 3)

paths <- routes %E>%
  as_tibble %>%
  filter(from > to) %>%
  mutate(passage = future_map2(from, to, ~passage(trans, sites[.x, ], sites[.y,], theta = 15)))

t1 <- paths %>% pull(passage) %>% brick %>% sum
plot(t1 / 300)

1:20 %>% accumulate(~.*(log(2) + 1), .init = t1) %>% .[[20]] %>% plot

feedback <- function(x) 

paths <- settlements %E>%
  group_by(from) %>% 
  top_n(-6, distance) %>%
  ungroup %N>%
  igraph::as.undirected(mode = 'collapse', edge.attr.comb = 'first') %>%
  as_tbl_graph
```


# getting rid of edge effects
```{r}
tst <- pts
tst[,2] <- tst[,2] + max(pts[,2]) - min(pts[,2]) * .5
tst[,1] <- tst[,1] + min(pts[,1])
plot(rbind(pts, tst))

min(pts[,1])
max(pts[,2])
```

